Structural Breaks in Inflation Dynamics within the European Monetary Union

### 2011-05-26 pkg/paper/Paper/draft.Rnw
updated paper and draft
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/glogis@103 b679b2f7-6539-45d0-a605-5ddcb976cde9

\Abstract{
  text when everything else finished, present innovation of paper
}

@@

Ever since the beginning of the European Union, the topic of a common currency
was a controversial issue. Although the Economic and Monetary Union (EMU) is now
a fact, the discussion about the economic effects of the euro is far from being
settled. The controversial topics range from the question of whether or not the
eurozone is indeed an optimal currency area (OCA) \fixme{Z: include a citation for
OCA or is this completely common knowledge?} all the way to the very
survival of the euro in light of the budgetary problems of some of its member
states. The effects of monetary unions on a number of macroeconomic indicators,
with inflation being the most important one in this context, is in the center of
an ongoing debate: this regards the question about short-run and steady-state
inflation uncertainty (in context to inflation expectations) dealt with in
\citet{cap}, or the structural convergence of the inflation rates in EU
countries, which is the topic of an investigation by \citet{palomba}.

The way towards the EMU basically consisted of three stages: stage~I
(1990--1994) as a phase of liberalization, stage~II (1995--1998) a phase of
convergence and stage~III (1999--2001) the transition period, which ended with
the introduction of the euro as legal tender. A possible fourth stage is the
continuing integration of new member states, mostly former communist countries.
Any new state that wants to join the EMU is first obliged to fulfill the
Maastricht criteria, which, besides setting rules for government debts and
interest rates, also requires the participation in the ERM~II (Exchange Rate
Mechanism~II) for two years, while the exchange rates towards the euro are not
allowed to cross the nominal band.

\cite{emerson} put emphasize on the fact that a high inflation rate is also more
variable and uncertain and thus causes more relative price variability, leading
to a less efficient price mechanism.  Failing to stabilize inflation leads to
severe economic problems. Throughout the literature, there is still a
considerable degree of uncertainty as to what extent the introduction of the
euro, or monetary unions in general, affect the inflation rate. The EMU requires
a country to possibly alter a number of economic policies and institutions,
which may demand for significant short term adjustments to take place.

@@

However, a deeper inquiry into this problem is in need of a method able to
capture changes in inflation dynamics. We contribute to this debate by
developing a new method for testing for structural breaks in the dynamics of the
inflation rates and thereby upon the effect of the EMU on a number of European
countries both within and outside the eurozone. The framework developed in this
paper makes it possible to test for any significant changes of a countries
inflation dynamics and thus helps interpreting the effects of the EMU. We focus
on changes in the first three moments of these inflation rates, which we assume
to be distributed according to a generalized logistic distribution (GL). The GL
distribution is a suitable one due to its flexibility to capture nonsymmetry and
because of its fatter tails. For this purpose, we develop a new estimation
approach building on previous research by \citet{Zeileis+Hornik:2007}.

This \fixme{Jesus: sell method better, why this one and not another} framework
allows us to assess the evolution of mean, variance and skewness over time. The
skewness parameter turns out to be quite important to measure changes in the
long run inflation dynamics, as it changes considerably over the time of the
analysis. This trend would be all but invisible using a normal distribution.  
Our estimation technique enables the simultaneous estimation of breaks in the
mean (which is important in the context of the Maastricht criteria), the
variance (which should be low for the price mechanism to be efficient) and the
skewness (where a change in skewness is evidence of a shift of the distribution
towards a new ``center of gravity'') of the inflation rates of interest. This
than can be used to search for answers regarding the question about the effect
of the EMU on the inflation dynamics of its member states or more general to
investigate the effects of external shocks.

@@

The type~I GL distribution (also known as skew-logistic distribution), as defined in \citet{johnson},
has the log-density function
%
\begin{eqnarray}
%% f(y ~|~ \theta, \sigma, \delta) & = &
%% \frac{\frac{\delta}{\sigma} \cdot \exp\{-\frac{y - \theta}{\sigma}\}}{(1+\exp\{-\frac{y - \theta}{\sigma}\})^{(\delta+1)}} 
\ell(y ~|~ \theta, \sigma, \delta) & = & \log(\delta) - \log(\sigma) - \frac{y - \theta}{\sigma} -
                                         (\delta + 1) \cdot \log\left(1 + \exp\left\{-\frac{y - \theta}{\sigma}\right\}\right)
\label{eq:dglogis}
\end{eqnarray}
%
with location~$\theta$, scale~$\sigma > 0$ and shape~$\delta > 0$. For $\delta = 1$ the
distribution simplifies to the logistic distribution, for $\delta < 1$ or $> 1$ it is
left- or right-skewed, respectively. The corresponding first three moments are
%
\begin{eqnarray}
\mathrm{E}(y) & = & \theta + \sigma (\gamma(\delta) - \gamma(1)) \\
\mathrm{Var}(y)  & = & \sigma^2(\gamma'(\delta)+\gamma'(1)) \\
\mathrm{Skew}(y) & = & \frac{\gamma''(\delta)-\gamma''(1)}{(\gamma'(\delta)+\gamma'(1))^{3/2}}
\end{eqnarray}
%
where $\gamma(\cdot)$, $\gamma'(\cdot)$, and $\gamma''(\cdot)$ are the digamma function and its
first and second derivative, respectively.

@@

Assuming a single stable set of parmaters (\ref{eq:h0}) is clearly not valid, as the
sequence of $\mathit{LM}$~statistics (see right panel of Figure~\ref{fig:slovenia1}) clearly
exceeds its 5\% critical value (red horizontal line), leading to a highly significant
$p$-value \Sexpr{spval[1]}. If a quadratic spectral kernal HAC estimator \citep{Andrews:1991}
would have been used to adjust (\ref{eq:efp}), the resulting $p$-value would be somewhat
larger with \Sexpr{spval[2]} but still clearly significant.

@@

This decrease in inflation is also conveyed by Figure~\ref{fig:slovenia2} that shows the
observed histograms of inflation rates before and after the breakpoint along with the
fitted GL distribution. This highlights that along with the decrease in mean level, the variance increases
somewhat and -- probably more interestingly -- the distribution changes from being right-skewed to
slightly left-skewed. This confirms that very high inflation rates that occurred in several
months before the breakpoint could be avoided afterwards. Furthermore, Figure~\ref{fig:slovenia2}
shows that the GL distribution fits the observed data very well, both before and after the
breakpoint, with $p$~values from $\chi^2$ goodness-of-fit tests of
\Sexpr{format(round(summary(x$model[[1]])$chisq.test$p.value, digits = 3), nsmall = 3)} and
\Sexpr{format(round(summary(x$model[[2]])$chisq.test$p.value, digits = 3), nsmall = 3)}, respectively.

@@

If there is only a single segment, the sup$\mathit{LM}$ test was non-significant at 5\%
level. In significant cases, the number of breakpoints was selected via the LWZ criterion
and subsequently all parameters (breakpoints and segment-specific GL parameters) were
estimated by ML. No HAC~correction was employed as there was only negligible autocorrelation
in the empirical scores after inclusion of the breakpoints (if any).

@@

\fixme{Z: I have reordered the groups in Table~1 which tries to be roughly in the order
of joining EU/ERM/EMU. Probably the following interpretations should start with a
very short overview of the groupings.}

The first group that draws particular attention is a subgroup of the PIIGS countries
(Portugal, Ireland, Italy, Greece, Spain) consisting of Ireland and
Spain. Both show a break in the summer of 2008 following the financial crisis
and the severe problems in both countries concerning the real estate situation.
In both countries this had a strong deflationary impact, with Ireland actually
showing negative inflation rates for an extended period of time; in Spain, which
was affected by a housing bubble of roughly the same size as Ireland, the
housing prices did not adjust so abruptly and in such a degree. This
deflationary trend, which for a time affected all countries, can be seen in
Figure~\ref{fig:glmean}, which depicts a rolling mean estimation, where the mean
is calculated using the first moment of a fitted GL distribution. There is
evidence of a decline in mean inflation, which returns to its previous path
roughly a year later.

The second group consists of the Eastern European countries: Czech Republic,
Estonia, Hungary, Poland, Slovakia and Slovenia. In almost all of these
countries, the mean and the variance of their inflation rates declined in the
later part of the 90s (Slovenia is an exception with an increase in variance
after the break). Most of them experienced a break in 1997/1998, with Poland,
Slovakia and Slovenia somewhat later. The break during the late 90s is a result
of these countries efforts to curb in inflation by decreasing money supply and
indicates the time when these countrie's actually overcame the biggest
transitionally shocks on their way towards free market economies. Slovakia is
very similar to Slovenia, as its inflation dynamics too changed roughly one year
prior to its entry into ERM~II.

Another interesting group consists -- with the exception of Austria and
Luxembourg -- of the countries of the former DEM-zone. All these countries have a
very low inflation rate in both mean and variance and their inflation dynamics
did not change.

The next group consists of two countries that decided not to be part of any
European Monetary System\footnote{If we ignore the short participating time of
both countries in the EMS.}, the United Kingdom and Sweden. Both show a clear
break in the early 90s. They can be traced back to economic crisis, the currency
crisis of the United Kingdom cumulating into the ``Black Wednesday'' in 1992 and
the run on the Swedish currency in the same year.

Austria, Finland and France adjusted to the requirements of the EMU during the
first phase of convergence. These are also the countries with the lowest
inflation rates in both mean and variance after the adjustment.

The last group consists of three PIIGS countries and Luxembourg. Greece and
Portugal adjusted to the ERM shortly before or afterwards its introduction.
Italy was successfull in reducing its inflation rate, whereas Luxembourg
experienced a doubling of its former -- very low -- inflation rate with the
variance rising by a factor of 14 with a strong shift in skewness towards the
left.

@@

The disparity of the national inflation rates is still an issue. Many researchers, like \citet{hofmann}  find a considerable amount of inflation
differentials. \citet{cap} estimate short-run and steady-state inflation uncertainty in 12 EMU countries and find a considerable degree of heterogeneity
across EMU countries in terms of average inflation and its degree of persistence. In a paper examining structural convergence of inflation rates in EU
countries, \citet{palomba} try to answer the question if during the 1990s the inflation rate dynamics of EU countries become more similar. They find that
convergence over time of inflation dynamics was only partly observable.  In a paper studying core inflation and using an aggregated euro area inflation
rate, \citet{morana} finds three regimes (roughly 1980--1984, 1984--1993 and 1993--2000) governing the core inflation rate.

Concerning inflation differentials, Figure~\ref{fig:glmean} depicts the inflation rates and a rolling mean estimation (with a window size of 12 months),
where the mean is calculated using the first moment of a fitted GL distribution.\footnote{If more countries have the same inflation rate in a given month,
the lines are overlapped and thus get darker. We wold see one black line if all countries would show the same month-on-month increase at any one time. The
red lines depict the means.} We do see that many of these countries follow a rather similar path. One event clearly visible is the financial crises and the
short deflationary phase starting somewhere around the autumn of 2007 and lasting a few months. Over time, there are hints, that the inflation rates
converged towards some common -- albeit broad -- mean level. For the inflation series of the whole EMU, as published by the \citet{oecd}, we see
an increase in mean from
$\Sexpr{round(eu_mom[1, 1], digits = 3)}$ (\Sexpr{start(eu)}--\Sexpr{eu_breaks[1]}) to
$\Sexpr{round(eu_mom[2, 1], digits = 3)}$ (\Sexpr{eu_breaks[1] + 1/12}--\Sexpr{eu_breaks[2]}) with a strong decrease to
$\Sexpr{round(eu_mom[3, 1], digits = 3)}$ (\Sexpr{eu_breaks[2] + 1/12}--\Sexpr{end(eu)}) following the financial crises.
Whereas variance has increased over time from
$\Sexpr{round(eu_mom[1, 2], digits = 3)}$ over $\Sexpr{round(eu_mom[2, 2], digits = 3)}$ to $\Sexpr{round(eu_mom[3, 2], digits = 3)}$.
Regarding skewness, we see a clear change towards a negative skewness from
$\Sexpr{round(eu_mom[1, 3], digits = 3)}$ over $\Sexpr{round(eu_mom[2, 3], digits = 3)}$ to $\Sexpr{round(eu_mom[3, 3], digits = 3)}$.
\fixme{Z: interpretation needs to be adapted to modified analysis}

@@

Regarding the general trend, Table~\ref{tab:results2} presents lower quartile,
median, and upper quartile (25\%, 50\%, 75\% quantiles) of the first three moments of the
inflation series in January \Sexpr{yrs[1]}, \Sexpr{yrs[2]}, \Sexpr{yrs[3]} and \Sexpr{yrs[4]}. The countries are grouped
into the ``core'' of the eurozone (Austria, Belgium, Finland, France, Germany,
Luxembourg and Netherlands), the PIIGS countries (Portugal, Italy, Ireland,
Greece and Spain) and all the others. For this, we took account of possible
breaks in the moments.

@@

In the core-zone, we see a decrease in the lower quartile from \Sexpr{yrs[1]} to
\Sexpr{yrs[4]}, with no change in the median inflation rate, which is due to the
fact, that there is no break after \Sexpr{yrs[2]} for these countries; however,
the interquartile range decreased by almost 20\%. The variance increased
slightly (but with a 15\% decrease in interquartile range), whereas skewness
stayed symmetric in the median, with almost no change in the interquartile
range.

The other countries were successfull in decreasing all three moments, but they
are still higher than the euro-zone countries. Most importantly, the
interquartile range for all moments decreased considerably showing a clear
convergence towards the core-zone countries.  The PIIGS countries decreased
inflation rates as well with a strong change of skewness towards more negative
values. The interquartile range of their inflation variance increased
considerably, which is mostly due to Spain and Ireland and their strong reaction
to the financial crisis.

@@

%Contrary to economic theory, which unfortunately does not provide any clear foundation for a proper answer to the question of whether or not the creation of a monetary union between two or more states is likely to reduce or increase the variability or even the level of the inflation rate, we were able to circumvent these problems by looking directly at the data and testing for breaks in the inflation rates dynamics.

\fixme{Jesus: this belongs to the intro section} An interesting approach to this question -- using a theoretical framework -- is taken by \citet{holte}. What he finds out via simulations of  different interest rate rules is that the standard deviation of the home CPI inflation rate can be substantially reduced by joining a monetary union. The effects of joining a monetary union on inflation variability in his framework depend on structural parameters like risk aversion, price flexibility, export demand elasticity, openness and shock correlations. However, due to the fact that not all of these parameters are known and that their interaction as well has to be estimated, the whole model is very dependent on a variety of assumptions.

%\fixme{Jesus: maybe, but you don't show that in your analysis} Regarding our results, it is safe to say, that the Eastern European countries were greatly successfull in decreasing inflation in both mean and variance. This is due to the fact that two of the most important inflationary channels, the money supply and the degree of openness, had to be arranged according to the criteria necessary for EMU entry. To achieve convergence in inflation rates, the growth in money supply had to be in concert with the M3 growth rate as decided by the European Central Bank (ECB). The opening of the economies made cheaper goods available and thus contributed to lower prices. Like the original countries participating in the ERM, the Eastern European countries had to synchronize money supply growth and interest rates in order to fix exchanges rates or otherwise risk the chance of severe disequilibria.

Any new country willing to join must first participate in the second exchange rate mechanism (ERM~II) with their currencies floating in a rather narrow band against the euro, which basically requires the same synchronization tools. 
For the countries in the core of the eurozone, the most disturbing trend seems to be the rise in inflation volatility in almost half of the euro countries. A very volatile inflation rate will likely  contribute to greater macroeconomic instability. Although \cite{grier} find no evidence that higher inflation uncertainty raises the average inflation rate -- at least in the case of the USA --  \cite{jaro} finds a positive correlation between the level and the standard deviation of inflation. This finding might be still more problematic if we take into account that on average the skewness parameter has decreased substantially in most of the euro-zone countries, and especially the PIIGS, indicating that in future higher inflation rates are more likely. This will put pressure on the ECB to achieve its inflation target of 2\%.

@@

In addition to Germany, Portugal, Belgium, Italy, Luxembourg and Spain, there
are a number of other countries, where the sales periods are regulated by law.
France has an individual time span regulated by each department. This should
pose no problem, since there is no global state level solution. In Greece there
are also two set sales-periods in winter and summer. In Italy the sales periods
are fixed each year by the Chambers of Commerce.  Portugal has two set sales
periods as well, but for two month (January and February as well as Mid-July to
Mid-September). The source for this information is \citet{konseur}. According to
\cite{pltm}, the official dates of the inclusion of these sales periods are:
January 1998 for Germany and Portugal, January 2000 for Belgium and January 1999
for Luxembourg and Ireland, while Italy and Spain introduced them in January
2001.

>>>>

\Abstract{
  In this paper, we test for structural breaks in the inflation dynamics of 21 European countries. To capture
  changes in the first three moments of the inflation rates, we assumed them to be distributed according to 
  a generalized logistic distribution, which we implemented in a new \proglang{R}~\Sexpr{paste(R.Version()[6:7], collapse = ".")} 
  \citep{R} package called \pkg{glogis}~\Sexpr{packageDescription("glogis")["Version"]}. We find breaks in all but 4 countries.
  Overall, there is a convergence towards a lower inflation rate, but with an increase in variance and a change towards negative skewness.
}

@@

Ever since the beginning of the European Union, the topic of a common currency
was a controversial issue. Although the Economic and Monetary Union (EMU) is now
a fact, the discussion about the economic effects of the euro is far from being
settled. The controversial topics range from the question of whether or not the
eurozone is indeed an optimal currency area (as developed in: \citet{Mundell:1961})\footnote{
For an overview, the interested reader is referred to \citet{Dellas:2009}}, 
all the way to the very survival of the euro in light of the budgetary problems of some of its member
states. The effects of monetary unions on a number of macroeconomic indicators,
with inflation being the most important one in this context, is in the center of
an ongoing debate: this regards the question about short-run and steady-state
inflation uncertainty (in context to inflation expectations) dealt with in
\citet{cap}, or the structural convergence of the inflation rates in EU
countries, which is the topic of an investigation by \citet{palomba}.

\readme{perhaps skip this paragraph?} The way towards the EMU basically consisted of three stages: stage~I
(1990--1994) as a phase of liberalization, stage~II (1995--1998) a phase of
convergence and stage~III (1999--2001) the transition period, which ended with
the introduction of the euro as legal tender. A possible fourth stage is the
continuing integration of new member states, mostly former communist countries.
Any new state that wants to join the EMU is first obliged to fulfill the
Maastricht criteria, which, besides setting rules for government debts and
interest rates, also requires the participation in the ERM~II (Exchange Rate
Mechanism~II) for two years, while the exchange rates towards the euro are not
allowed to cross the nominal band.

\clearpage
\cite{emerson} put emphasize on the fact that a high inflation rate is also more
variable and uncertain and thus causes more relative price variability, leading
to a less efficient price mechanism.  Failing to stabilize inflation leads to
severe economic problems. Throughout the literature, there is still a
considerable degree of uncertainty as to what extent the introduction of the
euro, or monetary unions in general, affect the inflation rate. The EMU requires
a country to possibly alter a number of economic policies and institutions,
which may require significant short term adjustments.

@@

However, a deeper inquiry into this problem is in need of a method able to
capture changes in inflation dynamics. We contribute to this  by
developing a new method for testing for structural breaks in the dynamics of the
inflation rates and thereby upon the effect of the EMU on a number of European
countries. The framework developed in this paper makes it possible to test for any 
significant changes of a countries inflation dynamics and thus helps interpreting 
the effects of the EMU. We focus on changes in the first three moments of these 
inflation rates, which we assume to be distributed according to a generalized logistic 
distribution (GL). The GL distribution is a suitable one due to its flexibility to 
capture nonsymmetry and because of its fatter tails. For this purpose, we develop a 
new estimation approach building on previous research by \citet{Zeileis+Hornik:2007}.

This \fixme{Jesus: sell method better, why this one and not another} framework
allows us to assess the evolution of mean, variance and skewness over time. The
skewness parameter turns out to be quite important to measure alterations in the
long run inflation dynamics, as it changes considerably over the time of the
analysis. This trend would be all but invisible using a normal distribution.  
Our estimation technique enables the simultaneous estimation of breaks in the
mean (which is important in the context of the Maastricht criteria), the
variance (which should be low for the price mechanism to be efficient) and the
skewness (where a change in skewness is evidence of a shift of the distribution
towards a new ``center of gravity'') of the inflation rates of interest. This
than can be used to search for answers regarding the question about the effect
of the EMU on the inflation dynamics of its member states or more general to
investigate the effects of external shocks.

@@

The GL distribution in the parameterization used here is defined in \citet{Abberger:2002}
% The type~I GL distribution (also known as skew-logistic distribution), as defined in \citet{Abberger:2002}
%\citet{johnson},
and has the log-density function\footnote{Which is similar to the type~1 GL distribution as defined in \citet{johnson}.}
%
\begin{eqnarray}
%% f(y ~|~ \theta, \sigma, \delta) & = &
%% \frac{\frac{\delta}{\sigma} \cdot \exp\{-\frac{y - \theta}{\sigma}\}}{(1+\exp\{-\frac{y - \theta}{\sigma}\})^{(\delta+1)}} 
\ell(y ~|~ \theta, \sigma, \delta) & = & \log(\delta) - \log(\sigma) - \frac{y - \theta}{\sigma} -
                                         (\delta + 1) \cdot \log\left(1 + \exp\left\{-\frac{y - \theta}{\sigma}\right\}\right)
\label{eq:dglogis}
\end{eqnarray}
%
with location~$\theta$, scale~$\sigma > 0$ and shape~$\delta > 0$. For $\delta = 1$ the
distribution simplifies to the logistic distribution, for $\delta < 1$ or $> 1$ it is
left- or right-skewed, respectively. The corresponding first three moments are
%
\begin{eqnarray}
\mathrm{E}(y) & = & \theta + \sigma (\gamma(\delta) - \gamma(1)) \\
\mathrm{Var}(y)  & = & \sigma^2(\gamma'(\delta)+\gamma'(1)) \\
\mathrm{Skew}(y) & = & \frac{\gamma''(\delta)-\gamma''(1)}{(\gamma'(\delta)+\gamma'(1))^{3/2}}
\end{eqnarray}
%
where $\gamma(\cdot)$, $\gamma'(\cdot)$, and $\gamma''(\cdot)$ are the digamma function and its
first and second derivative, respectively.

@@

Assuming a single stable set of parmaters (\ref{eq:h0}) is  not valid, as the
sequence of $\mathit{LM}$~statistics (see right panel of Figure~\ref{fig:slovenia1}) clearly
exceeds its 5\% critical value (red horizontal line), leading to a highly significant
$p$-value \Sexpr{spval[1]}. If a quadratic spectral kernal HAC estimator \citep{Andrews:1991}
would have been used to adjust (\ref{eq:efp}), the resulting $p$-value would be somewhat
larger with \Sexpr{spval[2]} but still clearly significant.

@@

\clearpage
This decrease in inflation is also conveyed by Figure~\ref{fig:slovenia2} that shows the
observed histograms of inflation rates before and after the breakpoint along with the
fitted GL distribution. This highlights that along with the decrease in mean level, the variance increases
somewhat and -- probably more interestingly -- the distribution changes from being right-skewed to
slightly left-skewed. This confirms that very high inflation rates that occurred in several
months before the breakpoint could be avoided afterwards. Furthermore, Figure~\ref{fig:slovenia2}
shows that the GL distribution fits the observed data very well, both before and after the
breakpoint, with $p$~values from $\chi^2$ goodness-of-fit tests of
\Sexpr{format(round(summary(x$model[[1]])$chisq.test$p.value, digits = 3), nsmall = 3)} and
\Sexpr{format(round(summary(x$model[[2]])$chisq.test$p.value, digits = 3), nsmall = 3)}, respectively.

@@

If there is only a single segment, the sup$\mathit{LM}$ test was non-significant at the 5\%
level. In significant cases, the number of breakpoints was selected via the LWZ criterion
and subsequently all parameters (breakpoints and segment-specific GL parameters) were
estimated by ML. No HAC~correction was employed as there was only negligible autocorrelation
in the empirical scores after inclusion of the breakpoints (if any).

@@

For Table~\ref{tab:results1}, we grouped the countries according to either their
similar timing of the changes in inflation dynamics or according to economic similarities, 
as in the case of the former communist countries in Eastern Europe.

The first group that draws particular attention consists -- with the exception of Austria and
Luxembourg -- of the countries of the former DEM-zone. All these countries had and still have 
a very low inflation rate in terms both of mean and variance and their inflation dynamics
did not change. This does make economic sense since the ECB is build after the tradition of
the Deutsche Bundesbank and the ECB did not -- until recently -- much diverge from the 
monetary policy of the former.

Austria, Finland and France adjusted to the requirements of the EMU during the
first phase of convergence. These are also the countries with the lowest
inflation rates in both mean and variance after the adjustment. In these countries,
the result of the adjustment process was a lower inflation rate and an overall
decrease in the variance.

The next group consists of three of the PIIGS (Portugal, Ireland, Italy, 
Greece and Spain) countries and Luxembourg. Greece and
Portugal adjusted to the ERM shortly before or after its introduction.
Italy was successfull in reducing its inflation rate, whereas Luxembourg
experienced a doubling of its former -- very low -- inflation rate with the
variance rising by a factor of 14 with a strong shift in skewness towards the
left. It is possible that in the case of Luxembourg, this still reflects the
adjustment due to the treatment of sales periods, even after seasonal adjustment
of the original series.

The forth subgroup consists of two of the PIIGS  countries, namely Ireland and Spain.  
Both show a break in the summer of  2008  following the financial crisis and the severe 
problems in both countries concerning   the real estate situation. In both countries this  
had a strong deflationary impact,  with  Ireland actually showing negative inflation rates 
for an extended period of time;   in Spain, which was affected by a housing bubble of  
roughly the same size as Ireland,  the housing prices did not adjust so abruptly and  
in such a degree. This deflationary  trend, which for a time affected all countries,  
can be seen in Figure~\ref{fig:glmean},  which depicts a rolling mean estimation,  
where the mean is calculated using the first  moment of a fitted GL distribution.  
There is evidence of a decline in mean inflation,  which returns to its previous 
path roughly a year later.

\clearpage
The next group consists of two countries that decided not to be part of the
European Monetary System\footnote{If we ignore the short participating time of
both countries in the EMS.}, the United Kingdom and Sweden. Both show a clear
break in the early 90s. These can be traced back to economic crisis, the currency
crisis of the United Kingdom cumulating into the ``Black Wednesday'' in 1992 and
the banking crisis in Sweden during the early 90s.

The last group consists of the Eastern European countries: Czech Republic,
Estonia, Hungary, Poland, Slovakia and Slovenia. In almost all of these
countries, the mean and the variance of their inflation rates declined in the
later part of the 90s (Slovenia is an exception with an increase in variance
after the break). Most of them experienced a break in 1997/1998, with Poland,
Slovakia and Slovenia somewhat later. The break during the late 90s is a result
of these countries efforts to curb in inflation by decreasing the growth of the money 
supply and indicates the time when these countries actually overcame the biggest
transitionally shocks on their way towards free market economies. Slovakia is
very similar to Slovenia, as its inflation dynamics too changed roughly one year
prior to its entry into ERM~II.

@@

The disparity of the national inflation rates within the EMU is still an issue. Many researchers, like \citet{hofmann}  find a considerable amount of inflation
differentials. \citet{cap} estimate short-run and steady-state inflation uncertainty in 12 EMU countries and find a considerable degree of heterogeneity
across EMU countries in terms of average inflation and its degree of persistence. In a paper examining structural convergence of inflation rates in EU
countries, \citet{palomba} try to answer the question if during the 1990s the inflation rate dynamics of EU countries become more similar. They find that
convergence over time of inflation dynamics was only partly observable.  In a paper studying core inflation and using an aggregated euro area inflation
rate, \citet{morana} finds three regimes (roughly 1980--1984, 1984--1993 and 1993--2000) governing the core inflation rate.

Concerning inflation differentials, Figure~\ref{fig:glmean} depicts the inflation rates and a rolling mean estimation (with a window size of 12 months),
where the mean is calculated using the first moment of a fitted GL distribution.\footnote{If more countries have the same inflation rate in a given month,
the lines are overlapped and thus get darker. We wold see one black line if all countries would show the same month-on-month increase at any one time. The
red lines depict the means.} We do see that many of these countries follow a rather similar path. One event clearly visible is the financial crises and the
short deflationary phase starting somewhere around the autumn of 2007 and lasting a few months. The general trend also suggests, that the inflation rates
converged towards some common -- albeit broad -- mean level. For the inflation series of the whole EMU, as published by the \citet{oecd}, we see
an increase in mean from
$\Sexpr{round(eu_mom[1, 1], digits = 3)}$ (\Sexpr{start(eu)}--\Sexpr{eu_breaks[1]}) to
$\Sexpr{round(eu_mom[2, 1], digits = 3)}$ (\Sexpr{eu_breaks[1] + 1/12}--\Sexpr{eu_breaks[2]}) with a slight decrease to
$\Sexpr{round(eu_mom[3, 1], digits = 3)}$ (\Sexpr{eu_breaks[2] + 1/12}--\Sexpr{end(eu)}) following the financial crises.
Whereas variance has increased considerably over time from
$\Sexpr{round(eu_mom[1, 2], digits = 3)}$ over $\Sexpr{round(eu_mom[2, 2], digits = 3)}$ to $\Sexpr{round(eu_mom[3, 2], digits = 3)}$.
Regarding skewness, we witness a movement towards negative skewness from
$\Sexpr{round(eu_mom[1, 3], digits = 3)}$ over $\Sexpr{round(eu_mom[2, 3], digits = 3)}$ to $\Sexpr{round(eu_mom[3, 3], digits = 3)}$.

@@

Regarding the general trend, Table~\ref{tab:results2} presents lower quartile,
median, and upper quartile (25\%, 50\%, 75\% percentiles) of the first three moments of the
inflation series in January \Sexpr{yrs[1]}, \Sexpr{yrs[2]}, \Sexpr{yrs[3]} and \Sexpr{yrs[4]}. The countries are grouped
into the ``core'' of the eurozone (Austria, Belgium, Finland, France, Germany,
Luxembourg and Netherlands), the PIIGS countries  and all the others. For this, we took account of possible
breaks in the moments.

@@

In the core-zone, we see a decrease in the lower quartile from \Sexpr{yrs[1]} to
\Sexpr{yrs[4]}, with no change in the median inflation rate, which is due to the
fact that there is no break after \Sexpr{yrs[2]} for these countries; however,
the interquartile range decreased by almost 20\%. The variance increased
slightly (but with a 15\% decrease in interquartile range), whereas skewness
stayed symmetric in the median, with almost no change in the interquartile
range.

The PIIGS countries decreased inflation rates as well with a strong change of skewness 
towards more negative values. The interquartile range of their inflation variance increased
considerably, which is mostly due to Spain and Ireland and their strong reaction
to the financial crisis.

The other countries were successfull in decreasing all three moments, but they
are still higher than the euro-zone countries. Most importantly, the
interquartile range for all moments decreased considerably showing a clear
convergence towards the core-zone countries.

@@

In contrast to \citet{holte}, whose simulations suggest that the standard deviation of the home 
CPI inflation rate can be substantially reduced by joining a monetary union, we fail to find a 
decrease in variance for the whole samples. However, this may be partly due to the financial crisis. 
For the countries in the core of the eurozone, the most disturbing trend seems to be the rise in 
inflation volatility in almost half of the euro countries. A very volatile inflation rate will 
likely  contribute to greater macroeconomic instability. Although \cite{grier} find no evidence 
that higher inflation uncertainty raises the average inflation rate -- at least in the case of the USA --
\cite{jaro} finds a positive correlation between the level and the standard deviation of inflation. 
This finding might be still more problematic if we take into account that on average the skewness 
parameter has decreased substantially in most of the euro-zone countries, and especially the PIIGS, 
indicating that in future higher inflation rates might be more likely. This will put pressure on the 
ECB to achieve its inflation target of 2\%. The Eastern European countries were greatly 
successfull in decreasing inflation in both mean and variance.

The results suggest that a GL distribution is appropriate for the modeling of the seasonally 
adjusted inflation rate series and that failing to account for a change in the skewness of the 
distribution might lead to wrong conclusions concerning the dynamics.

Another area for possible improvement is the data. It seems rather disturbing that adjustments 
in the computation of the HICP should lead to such considerable changes.

@@

In addition to Germany, Portugal, Belgium, Italy, Luxembourg and Spain, there
are a number of other countries where the sales periods are regulated by law.
In France the sale periods time span is regulated by each department. In Greece there
are also two set sales-periods in winter and summer. In Italy the sales periods
are fixed each year by the Chambers of Commerce.  Portugal has two set sales
periods as well, but for two month (January and February as well as Mid-July to
Mid-September). The source for this information is \citet{konseur}. According to
\cite{pltm}, the official dates of the inclusion of these sales periods are:
January 1998 for Germany and Portugal,  January 1999 for Luxembourg and Ireland and 
January 2000 for Belgium,  while Italy and Spain introduced them in January 2001.
